### 2018-01-29 includes/abstract.tex
a pass

    In this paper we present \fuzzer{}: an \smtlib{} problem fuzzer and generator for string solvers. We provide a rich repository of problems we generated with \fuzzer{} in \smt{} format. We also analyse \numSolvers{} mature string solvers: \theSolvers{}. We use \fuzzer{} to craft inputs that elicit poor performance in \us{} and \cvc{}, and analyse the causes of the performance degradations. We then describe an algorithm to mitigate the performance degradation in \us{}. Finally, we describe several bugs in \us{} that were exposed by \fuzzer{}.

>>>>

    In this paper, we introduce \fuzzer{}: a modular \smt{} problem transformer and generator for string solvers. We supply a repository of problems generated by \fuzzer{} in \smtfull{} format. We systemically compare \theSolvers{} on groups of such problems focusing on different combinations of string operators, and identify problems that are particularly challenging for some solvers. We briefly explain our observations and showcase how \fuzzer{} helped identify causes of the performance degradations in \us{}.

### 2018-01-29 includes/experiment.tex
Removing a section for now.
a pass

\section{Experimental Data}
\label{sec:data}

@@

    \begin{figure}[H]
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-extracts-small.png}
            \label{fig:concats-extracts-small}
            \caption{Performance on concats-extracts-small}
        \end{subfigure}
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/different-prefix.png}
            \label{fig:different-prefix}
            \caption{Performance on different-prefix}
        \end{subfigure}
        \caption{Problems hard for \cvc{}}
        \label{fig:cvc-hard}
    \end{figure}

    \begin{figure}[H]
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-balanced.png}
            \label{fig:concats-balanced}
            \caption{Performance on concats-balanced}
        \end{subfigure}
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-small.png}
            \label{fig:concats-small}
            \caption{Performance on concats-small}
        \end{subfigure}
        \caption{Problems hard for \us{}}
        \label{fig:z3str3-hard}
    \end{figure}

\section{Analyses}
\label{sec:analysis}

    In this section, we analyse our experimental results. We discuss the causes of \us{}'s poor performance on the two problem suites: \textit{concats-balanced} and \textit{concats-big}. We were not equipped well enough to debug the behaviours of the other solvers, but we provide our data in hopes that they will be useful to the solvers' authors.

    \subsection{\us{} on concats-balanced}

        \todo{Dmitry with help: Explain why \us{} performs poorly on this suite.}

@@

    \subsection{\us{} on concats-big}

        \todo{Dmitry with help: Explain why \us{} performs poorly on this suite.}

>>>>

\section{Experimental Data}
\label{sec:data}
\vspace{-0.1in}

@@

    \begin{figure}[h]
        \vspace{-0.2in}
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-extracts-small.png}                   
            \vspace{-0.25in}
            \caption{Performance on concats-extracts-small} 
            \label{fig:concats-extracts-small}
        \end{subfigure}
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/different-prefix.png}            
            \vspace{-0.25in}
            \caption{Performance on different-prefix}
            \label{fig:different-prefix}
        \end{subfigure}
        \vspace{-0.1in}
        \caption{Problems hard for \cvc{}}
        \label{fig:cvc-hard}
        \vspace{-0.3in}
    \end{figure}

    \begin{figure}[h]
    \vspace{-0.25in}
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-balanced.png}
            \label{fig:concats-balanced}
            \vspace{-0.25in}
            \caption{Performance on concats-balanced}
        \end{subfigure}
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-small.png}
            \label{fig:concats-small}
            \vspace{-0.25in}
            \caption{Performance on concats-small}
        \end{subfigure}
        \vspace{-0.1in}
        \caption{Problems hard for \us{}}
        \label{fig:z3str3-hard}
        \vspace{-0.3in}
    \end{figure}

\section{Usefulness to \us{}: A Case Study}
\vspace{-0.1in}

        We describe the defects and missing heuristics exposed by \fuzzer{} to illustrate why \fuzzer{} is a valuable asset in solver developments.

\vspace{-0.15in}
\subsection{Defects discovered and fixed in \us{}}
\vspace{-0.1in}

The first issue we found caused \us{} to perform poorly when equivalence classes 
of arithmetic terms were very large (e.g. the lengths of many strings were equal s
to 0). This issue was detected by the \texttt{Concats} problem suite. The slowdown 
was caused by a loop over terms of such equivalence class to search for an 
integer constant. The fix was to check the root of the equivalence class for the 
integer constant, as Z3 makes "interpreted terms" (e.g. constants) the root of 
equivalence classes whenever possible. This issue was last observed in commit 
\texttt{6308636}, and was fixed in commit \texttt{3865c45}.

The second issue, which was exposed by the \texttt{Lengths} suite, caused \us{} to 
perform poorly when a variable has a large length constraint. The slowdown was 
caused by the solver performing a linear search for a satisfying length when 
producing a model for the variable. This issue was an inconsistency between the 
implementation and the behavior described in the \us{} paper \cite{z3str3}. 
This issue was last observed in commit \texttt{66bc68f}, and was fixed in commit 
\texttt{7b536e9}.

\vspace{-0.15in}    
\subsection{Solving heuristics}
\vspace{-0.1in}

Discovering heuristics is crucial for solver efficacy improvements. Take 
conflicting constraint detection as an example. Solvers will eventually catch 
the offending terms, if they follow the string constraint solving algorithms and 
gradually propagate the (dis)equivalences discovered along the way. However, in 
many scenarios, a solver may be able to detect the conflicts before it tries to 
figure out the relationships among sub-terms and actually propagates the facts. 
The benefits are clear. It saves the efforts to reason about sub-terms and 
propagate the observations. As reported in existing works, heuristics usually 
lead to significant improvements. However, this is an extremely challenging 
process as the heuristic are usually learned from examples but it's usually hard 
to having multiple instances confirming such inefficiencies in the first place.

The instances in the \texttt{concats-big} generated by \fuzzer{} helped us 
discover a missing heuristic in \us{}. In particular, \us{} didn't make full use 
of the solving context (e.g. some terms are empty strings) to simplify the 
concatenations of a long list of string terms before trying to reason about the 
equivalences among sub-terms. It hence introduced a large number of unnecessary 
intermediate variables and fact propagations. We partially fixed this issue in 
our develop branch and observed some improvements.

% In this section, we analyse our experimental results. We discuss the causes of 
% \us{}'s poor performance on the two problem suites: \textit{concats-balanced} 
% and \textit{concats-big}. We were not equipped well enough to debug the 
% behaviours of the other solvers, but we provide our data in hopes that they will 
% be useful to the solvers' authors.

%     \subsection{\us{} on concats-balanced}
%         \todo{Dmitry with help: Explain why \us{} performs poorly on this suite.}

@@

    % \subsection{\us{} on concats-big}

    %     \todo{Dmitry with help: Explain why \us{} performs poorly on this suite.}

### 2018-01-29 includes/introduction.tex
Adding more sections. Adding references.
a pass

    Many SMT solvers that support the theory of strings have been developed over the past decade, and several have matured to the point of being performant and useful in practice. \theSolvers{}\cite{cvc4}\cite{z3str3} are \numSolvers{} such solvers. Moreover a standard input language for problem instances, \smtfull{}\cite{smt}, has been developed and widely adopted. To validate their solvers, many authors have relied on hand-crafted input suites or real-world examples from industrial applications \cite{cvc4-tests} \cite{z3str3-tests} \cite{z3str2-tests}. Indeed several suites of industrial and hand-crafted problems have been compiled \cite{kaluza} \cite{kausler} to both validate solvers and benchmark their performance. However, though these suites help debug solvers, they do not age well. As solvers continue to grow, fixed benchmarks leave more and more functionality untested.

    We introduce a modular fuzzer for string solvers, \fuzzer{}, and describe its utility as an exploratory validation tool. We show its effectiveness by presenting defects and limitations we were able to find in leading solvers.

@@

        \begin{enumerate}
            \item \textbf{The \fuzzer{} tool}: We describe a modular fuzzer which can transform and generate \smtfull{} string and regex problems. We briefly document its components and modular architecture. We provide some motivating usage patterns to demonstrate its usefulness as an exploratory tool for validating solvers. We additionally detail its regex-generating and transforming capabilities.
            \item \textbf{A repository of \smtfull{} problems}: We present a repository of \smtfull{} string and regex problem suites we generated with \fuzzer{}. We demonstrate the usefulness of these suites by revealing specific defects they helped uncover and fix in \us{}.
            \item \textbf{Experimental Results}: We present experimental runs of \theSolvers{} on the \fuzzer{} suites: \theSuites{}. We picked these suites because they uniquely made some solvers perform poorly, but not others.
            \item \textbf{Analysis}: We analyse our experimental results by inspecting the execution traces of the solvers on the given problems. For \us{} we pinpoint algorithmic limitations that cause the poor performance. For other solvers, we draw superficial conclusions based solely on the execution traces, and invite their authors to conduct further analysis.
        \end{enumerate}

>>>>

    Many SMT solvers supporting the theory of strings (e.g., CVC4 \cite{cvc4},  Z3str2 \cite{z3str2}, Norn \cite{norn} and Z3str3 \cite{z3str3}) have been developed over the past decade. To validate and benchmark their solvers, many authors have relied on hand-crafted input suites \cite{cvc4-tests} \cite{z3str3-tests} \cite{z3str2-tests} or real-world examples from industrial applications \cite{kaluza} \cite{kausler}. These test suites have helped developers identify implementation defects and develop more sophistic solving heuristics. Although extremely helpful, these suites can only cover a limited range of the problem space because they are either manually created or translated from carefully developed programs.

    Fuzzing, on the other hand, has already been widely used in similar scenarios to systematically generate more diverse test cases. Guided by different rules, it can also be applied to generate different instances that are particularly designed to challenge selected capacities. Inspired by traditional fuzzers, we introduce a modular fuzzer for string solvers, \fuzzer{}, and describe its utility as an exploratory validation tool. We demonstrate its effectiveness by presenting defects and limitations we were able to identify in leading solvers.
%     Moreover, a standard input language for problem instances, \smtfull{}\cite{smt}, has been developed and widely adopted.

@@

        \begin{enumerate}
            \item \textbf{The \fuzzer{} tool}: We describe a modular fuzzer which can transform and generate \smtfull{} string and regex problems. We briefly document its components and modular architecture. We provide example usages to demonstrate its usefulness as an exploratory tool for validating solvers. We additionally detail its regex-generating and transforming capabilities.
            \item \textbf{A repository of \smtfull{} problems}: We present a repository of \smtfull{} string and regex problem suites we generated using \fuzzer{}. We demonstrate the usefulness of these suites by revealing specific defects they helped uncover and fix in \us{}.
            \item \textbf{Experimental Results}: We compare \theSolvers{} on the \fuzzer{} suites and selectively present the experimental result on \theSuites{}. We highlighted these suites because they uniquely made some solvers perform poorly, but not others.
            \item \textbf{Analysis}: We analyze our experimental results by inspecting the execution traces of the solvers on the given problems. For \us{}, we pinpoint algorithmic limitations that cause the poor performance. For other solvers, we draw superficial conclusions based solely on the execution traces, and invite their authors to conduct further analysis.
        \end{enumerate}

### 2018-01-29 includes/problems.tex
Adding placeholder reference for Ifaz's paper. Removing some TODOs.
a pass

\section{Problem Suites}
\label{sec:problems}

    We generated several suites of string and regex problems using \fuzzer{}. \footnote{For review anonymity, we removed the link to the problem repository. It will be provided separately in the artifact submission track and will appear in the final version of the paper.} Table~\ref{tbl:generated} describes the \texttt{generated} suite that we generated using \generator{}.

    \begin{table}
        \centering
        \begin{tabular}{|l|l|}
            \hline
            \textbf{Name}             & \textbf{Problems have ...} \\ \hline
            \textit{Concats}          & A single, deep tree of concats. \\ \hline
            \textit{Concats-Extracts} & A single concat tree, and random character extractions. \\ \hline
            \textit{Lengths}          & A single, large length constraint on a variable. \\ \hline
            \textit{Lengths-Concats}  & Variables and their concats, each with fixed lengths. \\ \hline
            \textit{Overlaps}         & A formula of the form A.X = X.B. \\ \hline
            \textit{Regex}            & A test for regex membership of arbitrary complexity. \\ \hline
            \textit{Many-Regexes}     & A test for membership in many random regexes. \\ \hline
            \textit{Regex-Deep}       & A test for membership in a regex with many nested operators. \\ \hline
            \textit{Regex-Pair}       & A test for membership in one regex, but not another. \\ \hline
            \textit{Regex-Lengths}    & A test for membership in a regex, and a length constraint. \\ \hline
        \end{tabular}
        \caption{\generator{}-generated problem sets.}
        \label{tbl:generated}
    \end{table}

    The \texttt{transformed} suite contains benchmarks derived from existing seed problems by iteratively applying \transformer{}. Every problem in the suite is named according to its seed and the transformations it undertook. For example, a hypothetical file named ``seed-fuzz-graft.smt2'' began as a file named ``seed.smt'' and was transformed twice: first by fuzz, and then by graft. The suite is divided into two categories. For an overview see Table~\ref{tbl:transformed}.

    \begin{table}
        \centering
        \begin{tabular}{|l|l|r|}
            \hline
            \textbf{Name}   & \textbf{Seed}                            & \textbf{Quantity} \\ \hline
            \textit{Amazon} & Two industrial regex membership problems & 516\\ \hline
            \textit{Regex}  & Z3str2 regular expression test suite     & 10,577\\ \hline
        \end{tabular}
        \caption{\transformer{}-generated problem sets.}
        \label{tbl:transformed}
    \end{table}

    The first category, labeled Amazon, started as two seed files.
    The first seed, z3-regex-1.smt2, encodes the assertions ``S not in aaabbb*'' and ``S in aaa*bbb*'' for a string constant S,
    while the second seed, z3-regex-2-negative.smt2, encodes the assertions ``S in aaabbb*'' and ``S not in aaa*bbb*''.
    Both problems come from our industrial collaborators.
    In total, 516 unique problems were generated from these two seeds.

    The second category, Regex, is seeded by the z3str2 regular expression test suite \cite{z3str2-tests}.
    This test suite contains 42 problems dealing with a wide variety of regular expressions.
    Through cumulative transformations, as before, we were able to produce 10,577 unique benchmarks.

    \subsection{Usefulness to \us{}}

        This section describes the bugs we discovered in \us{} (that were subsequently fixed) by testing it with the \fuzzer{} problem suites.

        The first bug we found caused \us{} to perform poorly when equivalence classes of arithmetic terms was very large (e.g. where many strings were equal to the empty string, and thus had length 0). We discovered this bug using the \textit{Concats} problem suite. The slowdown was caused by a loop over the equivalence class of these terms to search for an integer constant. The fix was to check the root of the equivalence class for the integer constant, as Z3 makes "interpreted terms" (i.e. constants) the root of equivalence classes whenever possible. This bug was last observed in commit 6308636, and was fixed in commit 3865c45.

        The second bug we found caused \us{} to perform poorly when a variable had a large length constraint. We discovered this bug using the \textit{Lengths} problem suite. The slowdown was caused by the solver performing a linear search for a satisfying length when producing a model for the variable. This bug was an inconsistency between the implementation and the behaviour described in the \us{} paper \cite{z3str3}. This bug was last observed in commit 66bc68f, and was fixed in commit 7b536e9.

        \todo{Federico - Can we describe the floating point error we found?}

        \todo{Dmitry: maybe show graphs with before and after?}

>>>>

\section{Problem Suites}
\label{sec:problems}
\vspace{-0.1in}

\subsection{Generated suite}
\vspace{-0.1in}
    We generated several suites of string and regex problems using \fuzzer{}\footnote{For review anonymity, we removed the link to the source code. It will be provided separately in the artifact submission.}. Table~\ref{tbl:generated} describes the suite \texttt{generated} by \generator{}.

\begin{table}[h]
\vspace{-0.4in}
\centering
\caption{\generator{}-generated problem sets.}
\label{tbl:generated}
{
  \footnotesize
  \bgroup
  \def\arraystretch{1}
  \resizebox{0.8\columnwidth}{!}{
        \begin{tabular}{|l|l|}
            \hline
            \textbf{Name}             & \textbf{Problems have ...} \\ \hline
            \textit{Concats}          & A single, deep tree of concats. \\ \hline
            \textit{Concats-Extracts} & A single concat tree, and random character extractions. \\ \hline
            \textit{Lengths}          & A single, large length constraint on a variable. \\ \hline
            \textit{Lengths-Concats}  & Variables and their concats, each with fixed lengths. \\ \hline
            \textit{Overlaps}         & A formula of the form A.X = X.B. \\ \hline
            \textit{Regex}            & A test for regex membership of arbitrary complexity. \\ \hline
            \textit{Many-Regexes}     & A test for membership in many random regexes. \\ \hline
            \textit{Regex-Deep}       & A test for membership in a regex with many nested operators. \\ \hline
            \textit{Regex-Pair}       & A test for membership in one regex, but not another. \\ \hline
            \textit{Regex-Lengths}    & A test for membership in a regex, and a length constraint. \\ \hline
        \end{tabular}  
  }
  \egroup
  \vspace{-0.2in}
}
\end{table}

%     \begin{table}
%         \centering
%         \begin{tabular}{|l|l|}
%             \hline
%             \textbf{Name}             & \textbf{Problems have ...} \\ \hline
%             \textit{Concats}          & A single, deep tree of concats. \\ \hline
%             \textit{Concats-Extracts} & A single concat tree, and random character extractions. \\ \hline
%             \textit{Lengths}          & A single, large length constraint on a variable. \\ \hline
%             \textit{Lengths-Concats}  & Variables and their concats, each with fixed lengths. \\ \hline
%             \textit{Overlaps}         & A formula of the form A.X = X.B. \\ \hline
%             \textit{Regex}            & A test for regex membership of arbitrary complexity. \\ \hline
%             \textit{Many-Regexes}     & A test for membership in many random regexes. \\ \hline
%             \textit{Regex-Deep}       & A test for membership in a regex with many nested operators. \\ \hline
%             \textit{Regex-Pair}       & A test for membership in one regex, but not another. \\ \hline
%             \textit{Regex-Lengths}    & A test for membership in a regex, and a length constraint. \\ \hline
%         \end{tabular}
%         \caption{\generator{}-generated problem sets.}
%         \label{tbl:generated}
%     \end{table}

\vspace{-0.2in}
\subsection{Transformed suite}
\vspace{-0.1in}
    The \texttt{transformed} suite contains benchmarks derived from existing seed problems by iteratively applying \transformer{}. Table~\ref{tbl:transformed} shows the overview. Every problem in the suite is named according to its seed and the transformations it undertook. For example, input \texttt{seed-fuzz-graft.smt2} was transformed by applying \texttt{fuzz} and then \texttt{graft} from \texttt{seed.smt2}.

\begin{table}[h]
\vspace{-0.35in}
\centering
\caption{Transformed problem sets generated by \transformer{}.}
\label{tbl:transformed}
{
  \footnotesize
  \bgroup
  \def\arraystretch{1}
  \resizebox{0.6\columnwidth}{!}{
          \begin{tabular}{|l|l|r|}
            \hline
            \textbf{Name}   & \textbf{Seed}                            & \textbf{Quantity} \\ \hline
            \textit{Amazon} & Two industrial regex membership problems & 516\\ \hline
            \textit{Regex}  & Z3str2 regular expression test suite     & 10,577\\ \hline
        \end{tabular}
    }
  \egroup
  \vspace{-0.2in}
}
\end{table}

%     \begin{table}
%         \centering
%         \begin{tabular}{|l|l|r|}
%             \hline
%             \textbf{Name}   & \textbf{Seed}                            & \textbf{Quantity} \\ \hline
%             \textit{Amazon} & Two industrial regex membership problems & 516\\ \hline
%             \textit{Regex}  & Z3str2 regular expression test suite     & 10,577\\ \hline
%         \end{tabular}
%         \caption{\transformer{}-generated problem sets.}
%         \label{tbl:transformed}
%     \end{table}

    The Amazon category is obtained from two seeds: (1) \texttt{z3-regex-1.smt2} encodes the assertions \texttt{(S not in aaabbb*)} and \texttt{(S in aaa*bbb*)} for a string constant S; (2) \texttt{z3-regex-2-negative.smt2} includes regex \texttt{(S in aaabbb*)} and \texttt{(S not in aaa*bbb*)}.
    Both problems come from our industrial collaborators. In total, 516 unique problems were generated from these two seeds.

    The Regex category is seeded by the z3str2 regular expression test suite \cite{z3str2-tests}.
    This test suite contains 42 problems dealing with a wide variety of regular expressions.
    Through cumulative transformations, we have $10,577$ unique benchmarks.
